\documentclass{article}
\usepackage{iclr2026_conference}
\input{math_commands.tex}
\usepackage{booktabs}
\usepackage{array}
\usepackage{url}

\title{Latent Fixture Reasoning}
\author{Anonymous Authors}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Assembly robots often treat support relations and fixtures as visible scene
geometry or as constraints supplied before planning. This paper studies the
missing case: a part fails to move because an unobserved bolt, latch, adhesive
patch, or hidden stop is still constraining it. We propose \emph{latent fixture
reasoning}: infer a discrete hidden fixture cause from small manipulation
outcome residuals, then choose the release or reorientation action implied by
that cause. A bounded literature sweep over 1168 arXiv/Crossref
records suggests that the nearest work covers contact-rich manipulation,
constraint reasoning, tactile inference, active perception, and fixture-aware
assembly, but rarely treats hidden fixtures as an explicit action-selecting
state variable. In a deterministic toy benchmark, geometry-only manipulation
collides with many fixture modes, while latent fixture reasoning sharply
reduces wrong release actions when the hidden fixture belongs to the known
taxonomy. A v2 out-of-taxonomy stress shows the boundary: an unseen cable-tie
fixture causes forced known-taxonomy inference to fail in all 2,000 seeds, and
a guarded detector avoids nearly all damage only by abstaining. The evidence is
modest, but it exposes a useful state abstraction for robots that must learn
what is holding the world in place by acting on it.
\end{abstract}

\paragraph{V2 hardening note.}
This version narrows the claim. Latent fixture reasoning is not a universal
unknown-constraint solver. The mechanism is supported only when the fixture
taxonomy and release-action menu are calibrated, or when the robot is allowed
to abstain on out-of-distribution residuals.

\section{Introduction}
When a robot pulls on a part and it does not move, the failure is not just a
low-level control error. It can be evidence that something unseen is holding
the part in place. In assembly, repair, household manipulation, and field
robotics, that hidden cause may be a bolt, a latch, adhesive, a cable, a clamp,
or a geometric stop outside the camera view. A planner that represents only
visible geometry can repeatedly choose unsafe or unproductive actions.

We call this problem \emph{latent fixture reasoning}. The key object is an
interpretable hidden fixture variable. It is latent because the robot does not
observe the fixture directly. It is a fixture because it constrains motion. It
is reasoning, rather than only prediction, because the inferred cause changes
the next action: unbolt, unlatch, heat, reorient, or pull.

The paper makes three contributions. First, it separates latent fixture causes
from generic uncertainty in manipulation. Second, it defines a small posterior
interface from probe residuals to fixture hypotheses. Third, it provides a
minimal deterministic simulation showing why this state variable can reduce
wrong fixture actions.

\section{Problem}
Let $x$ denote the visible object state and let $z$ be an unobserved fixture
mode in a calibrated set
$\{\mathrm{free}, \mathrm{bolt}, \mathrm{latch}, \mathrm{adhesive},
\mathrm{stop}\}$. A robot chooses a probe action $a_p$ and observes an outcome
residual $o$: displacement, lift compliance, twist compliance, and thermal
softening response. The robot then chooses a task action $a_t$ from direct pull,
unbolt, unlatch, heat-and-peel, or reorient. The central question is not whether
the robot can predict the next state in general. It is whether $o$ identifies a
fixture cause well enough to avoid the wrong physical release action. If the
true fixture is outside the calibrated set, the safe target behavior is
abstention or additional diagnosis, not confident release.

\section{Mechanism}
The proposed controller maintains a small posterior over fixture modes,
\[
  p(z \mid o) \propto p(o \mid z)p(z).
\]
The likelihood model can be hand-written, learned, or calibrated from robot
data. In this paper we use a hand-written model only to isolate the mechanism.
The controller is allowed to perform a reversible probe before committing to a
release action. If confidence is low, it performs an additional reversible
probe rather than applying a high-force release action. In the hardened v2
stress, we also expose a simple distance-to-known-prototype guard: residuals
far from every known fixture prototype are treated as out-of-taxonomy and
should trigger abstention.

This differs from a generic world model in two ways. First, the latent variable
has a robotics meaning: it is a hidden support or fixture cause. Second, the
variable is directly coupled to an action menu. A high posterior on bolt selects
unbolt; a high posterior on adhesive selects heat-and-peel; a high posterior on
hidden stop selects reorientation.

\section{Evidence}
We implemented a deterministic toy benchmark with 2,000 seeds. Each seed
samples one fixture mode, generates noisy probe residuals, and evaluates three
policies. The geometry-only policy assumes visible geometry is complete. The
reactive policy first fails, then uses a single heuristic residual. The latent
fixture policy infers the fixture before choosing a release action.

\begin{table}[h]
\centering
\small
\begin{tabular}{lrrrrrr}
\toprule
Policy & Success & Coll. & Wrong & Steps & Return & Conf. \\
\midrule
Geometry only & 621 & 1379 & 1379 & 3.689 & -0.696 & 0.000 \\
Reactive & 1612 & 1767 & 388 & 5.758 & 0.332 & 0.207 \\
Latent fixture & 1995 & 5 & 5 & 7.077 & 0.871 & 0.941 \\
\bottomrule
\end{tabular}
\caption{Toy hidden-fixture benchmark. The mechanism is valuable when a
visible-geometry policy cannot distinguish free motion from a hidden fixture
constraint.}
\end{table}

The absolute numbers should not be read as benchmark performance. They show a
mechanism: when the hidden cause selects a different release action, estimating
that cause before committing can reduce physical mistakes.

\subsection{V2 out-of-taxonomy stress}
The hostile reviewer attack is taxonomy misspecification. The v1 controller
knows only free, bolt, latch, adhesive, and hidden-stop modes. We add a
``cable-tie'' fixture whose residual signature is outside that menu and whose
correct action is also absent from the release menu. A forced latent-fixture
classifier must map it to a known class, so every committed release is wrong.
The guarded variant checks distance from the nearest known prototype and can
avoid damage only by refusing to act. Table~\ref{tab:unknown-fixture-stress}
therefore reports a negative result, not a new success claim.

\input{docs/unknown_fixture_stress_table.tex}

\section{Related Work Boundary}
The literature sweep covers contact-rich manipulation, task-and-motion planning,
tactile inference, support-relation reasoning, robot world models, and assembly
fixture planning. These areas provide the hostile prior. The narrow boundary for
this paper is an interpretable latent fixture graph whose nodes are not merely
unobserved state, but action-selecting causes of constraint.

\section{Limitations}
The experiment is a toy model with hand-coded fixture types and synthetic
residuals. It does not prove that the abstraction scales to real assemblies,
and the v2 stress shows that the method is brittle when the true fixture class
or required release action is missing from the taxonomy. The next required step
is a physical or high-fidelity simulated benchmark where fixtures are hidden
from perception, revealed through controlled probes, and sometimes outside the
training taxonomy.

\section{Conclusion}
Latent fixture reasoning asks robots to treat failed motion as evidence about
what hidden support is still constraining the world. The contribution is a
small but useful shift in representation: from visible geometry plus generic
uncertainty to an interpretable hidden fixture cause that chooses the next safe
manipulation action.

\section*{References}
\small
The accompanying \texttt{docs/related\_work\_matrix.csv} contains the bounded recovery
literature sweep used for this mechanism note.

\end{document}
